# R - Introduction
## R - Basic Concepts and applications 
**R (programming language)** is a language and environment for statistical computing and graphics 

<-- logo image for Rstudio --> 

## Why R? 

1. **Developed by Statisticians** - R is Excellent for Statistical Computing and Analysis 
2. **Open-source (Free)** - R is an open-source programming language. Anyone can work with R without any license or fee. Due to this, R has a huge community that contributes to its environment. 
3. **A Large Variety of Specialized Libraries** - These libraries support and enhance the R development environment.
4. **R Studio** - A graphic user interface to R - easier to grasp and learn. 
5. **Cross-platform Support** - R is machine-independent. It supports the cross-platform operation. Thus, it is usable on many different operating systems.
6. **Can do Data Cleansing, Data Wrangling, and Web Scraping** - R can collect data from the internet through web scraping and other means. It can also perform data cleansing. R is also useful for data wrangling which is the process of converting raw data into the desired format for easier consumption. 
7. **Powerful Graphics** - R has extensive libraries that can produce production quality graphs and visualizations. 
8. **Compatible with other Programming Languages** - R is compatible with other languages like C, C++, and FORTRAN. Other languages like .NET, Java, Python can also directly manipulate objects.
9. Compatible with high performance computing and Big Data Analytics. 

## R studio
**Rstudio** is an integrated development environment(IDE) for R. 

It includes a console, syntax-highlighting editor that supports direct code execution as well as tools for plotting history, debugging and work space management. 
RStudio - installation link <-link in the first 7 minutes->

Outline the process of creating and running R script - a screenshot 
<--Check R Version --> 

### R - Look and Feel

<-- R screenshot image from the terminal-->

<-- default screenshot image for R Studio --> 
explain each section/quadrant
- text editor 
- stand input/output (console/terminal/jobs)
- all data and its properties quadrants 
- plots and file directory structure

## Basic Mathematics like operators 
Addition - the `+` operator is used to add two or more values 
```{r}
2+3
```

Subtraction - the `-` operator is used to subtract two numerical values 
```{r}
6-2
```

<-- add more content for operators-->

## R Markdown 
<-- Definition of R Markdown -->
<--Introduce how to create R markdown file using screenshots --> 


Each chunk can be run separately in a single file, also the variable in one chunk can be reused by the next/later chunks 

Anything added outside the chuck will not be executed in the program

The R markdown can be knitted to a html, pdf, or word file 
<--Include screenshots-->
