# Data Types and Data Structures
## Data types 
<-- add more explanations to each data types --> 

- Numeric
```{r}
a=23.5
class(a) #check the data type of a 
```

- Integers
```{r}
a=23L #add L to show it is an integer
class(a)
```

- Characters
```{r}
a="DNA"
class(a)
```

- Logical 
```{r}
a=TRUE #logical data types is either TRUE or FALSE only
class(a)
```

- Complex, 
```{r}
a=2+3i # Complex data types have 'i' at the end of each number
class(a)
```

- Raw
```{r}
a=charToRaw("DNA")
print(a)
class(a)

# convert back to character 
b=rawToChar(a)
class(b)
```

## Data Structures 
This is the organization of data into one or multiple data values in specific structures. Different types of data structures in R include;
- Vector
- Matrix
- Data frame

### Vector
A vector is a single entity consisting of a collection of things. Here is an example of a vector.
```{r}
marks = c(23, 67, 98, 34, 98, 21)
print(marks)
```

The `class` function is utilized to determine the data types present within vector data values.
```{r}
class(marks)
```
The vector "marks" consist of only numeric values 

`is.vector` function is used to check if the variable is a vector 
```{r}
is.vector(marks)
```

unlike matrix and data frame, vector has no dimension 
```{r}
dim(marks)
```

`length()` function is used to count number of elements in vectors
```{r}
length(marks)
```

Index is the position of an element in a vector, in R it starts at index 1 - lets say we find the third element by index 3
```{r}
marks[3]
```
value "98" is at index 3, or the third in the vector. 

Vectors can be sliced to obtain values over a range of indices. For instance the code below shows how to retrieve the from the second to the fourth values as a vector
```{r}
print(marks[2:4])
is.vector(marks[2:4]) # confirm if the retrieved values are in a vector
```

An element at a specific index in a vector can be excluded by adding a `-` sign before the index value. 
```{r}
marks[-2] #exclude the element at index 2
```

`rev()` command is used to reverse the order of elements in a vector
```{r}
rev(marks)
```

#### Mathematical Operations in a vector 
The summary/descriptive statistics are calculated by `summary()` command. 
```{r}
summary(marks)
```

`sum()`, `median()`, and `mean()` are used to calculate the total, median, average and the standard deviation of the values in a vector
```{r}
print("MARKS")
print(paste("TOTAL: ", sum(marks)))
print(paste("MEDIAN: ", median(marks)))
print(paste("AVERAGE: ", mean(marks)))
```

vector multiplication
```{r}
double_marks =2 * marks 
marks
double_marks
```

Vector division - divide the marks by 2
```{r}
half_marks = marks/2
half_marks
```

Multiply 2 vectors of equal length
```{r}
a = c(3, 5, 1)
b = c(7, 3, 9)
ab = a*b
ab
ba = b*a # is the same as ab
ba
```

Divide 2 vectors of equal length 
```{r}
# Divide vector a by b
abdiv=a/b
abdiv
# Divide vector b by a
badiv=b/a
badiv
```

However, when multiplying vectors of unequal length the shorter one is replicated to match the longer vector. It will then return a warning
```{r}
e=c(1,2,3,4,5)
f=c(1,2)
ef = e*f #it shows an error
ef #shows results since f is replicated to match e as f=c(1,2,1,2,1)
```

Multiple vectors can be concatenated/combined to come up with one giant vector 
```{r}
a
b
z=c(a,b,a) #concatenates the vectors 
z
```

#### Character Vectors 
Vectors can also contain character data types for instance 
```{r}
my_name = c("My", "name", "is", "Vipin")
my_name[5] = "Singh" #insert at the end 
my_name
```

Combining the vectors to a single string 
```{r}
print(paste(my_name, collapse=" "))
```

Calculate the summary/descriptive statistics of the vector by function `summary()`. It finds;
- Count/length
- Class (data type)
- Mode 
```{r}
summary(my_name)
```

#### Vectors with mixed data types 
A vector can also consist of characters values and numeric values for instance 
```{r}
numbers=c(1,"two", 3, "three")
numbers
```

however the numeric elements in the vector are recognized by R as character data type. They can be converted to numeric by; 
```{r}
as.numeric(numbers[1]) + 2
```

the integers can be converted by;
```{r}
as.integer(numbers[1])
```

#### Named Vectors 
Variable names can be assigned to vectors like;
```{r}
named_vector=c(EcoR1="GAATTC", HindIII="AAGCTT", Pst1="CTGCAG")
named_vector
```

to access the names of the values is;
```{r}
names(named_vector)
```

A vector element can be accessed using its name
```{r}
named_vector["EcoR1"] # find the value of a vector by its name 
```

#### Generating  number series as vectors 
The `seq` function in R is used to generate sequences of numbers. It takes several arguments, including `from`, `to`, `by`, and `length.out`, among others, to specify the range and increment of the sequence. Here's a brief overview of its usage:

- `from`: The starting value of the sequence.
- `to`: The end value of the sequence.
```{r}
# Generate a sequence from 1 to 10
series = seq(from=1, to=20)
series 

# It can also be written as
series = seq(1,20)
series
```

```{r}
class(series)
```

- `by`: The increment between consecutive values in the sequence.
```{r}
# generate numbers 0 to 10 incremented by 0.5
series3=seq(0, 10, by=0.5)
series3
```

- `length`: The desired length of the sequence.
```{r}
# generate 10 numbers from 0 to 6
series4=seq(0, 6, length=10)
series4
seq(0, 6)
```
- `along.with`: An optional vector argument specifying the length and names of the output sequence.
```{r}
# Generate a sequence along with a vector
seq(along.with = c("a", "b", "c"))
```

#### Null data points in vectors 
NA data (Not available or blank) for instance 
```{r}
marks=c(78,65, 98, 87, 89, NA)
sum(is.na(marks)) #Count the null values in a vector 
```

Other inbuilt functions for mathematical operations cannot be done if Null values exists in a vector unless they are removed/ignored 
```{r}
#sum(marks) #returns an error
sum(marks, na.rm = TRUE) #remove null values before calculating the sum
median(marks, na.rm = TRUE)
summary(marks, na.rm = TRUE)
```

### Matrix 
A matrix is a two dimensional data type that contain a single class of data. The code below shows one can produce a matrix from a vector 
```{r}
vector1 = seq(1, 9) 

# Convert to matrix
## create by column 
data1=matrix(vector1, ncol=3)
data1
```
A vector of values 1 to 9 is being converted to a matrix where the values are being arranged columnwise by default. 

A matrix has a multiple dimensions, the most common type of matrix is two dimesnional. 
```{r}
dim(data1)
```

`is.matrix()` function is used to confirm if a given variable is a matrix and it return a boolean value. 
```{r}
is.matrix(data1)
```

A matrix can also be created row-wise from a vector.
```{r}
## create by row 
data2=matrix(vector1, ncol=3, byrow=TRUE)
data2 # is a transpose of data1
```

Matrix is recognized either as a matrix or array by R
```{r}
class(data1) 
```

To access a specific data point in a matrix, the matrix is indexed by row then column for instance `matrix_data[row_index, column_index]`
```{r}
data1[3, 2]
```

To access a single row, in this case we find the second row which will be returned as a vector 
```{r}
data1
row2 = data1[2,] # access row 2 
is.vector(row2) #can be accessed by row 2
```

To access a single column 
```{r}
data1
col3=data1[,3] # access column 3
is.vector(col3) #can be accessed by column 3
```

Count the number of rows in a matrix 
```{r}
nrow(data1)
```
`data1` has 3 rows

Count the number of columns in a matrix
```{r}
ncol(data1)
```

#### Mathematical Operations in a matrix 
**Matrix Addition**
```{r}
data1
data2 = data1 + 3
data2
```
Adding value 3 to a matrix adds each value in the matrix by 3

**Matrix Subtraction**
```{r}
data3 = data1-1 #reduce each value by 1
data3
```
Subtracting 1 to `data1` subtract each value in the matrix by 1

**Matrix Multiplication(scalar)**
```{r}
data4 = data1*5
data4
```

**Matrix division**
```{r}
data5 = data1/2
data5
```


### Data frame
is a two dimensional data structure, like a 2d array/matrix with rows and columns. 

Lets convert a matrix into a dataframe 

```{r}
vector1 = c(1:12)
matrix1 = matrix(vector1, ncol=4) #create a matrix from the vector 

# Adding a column student 
Students=c("Pragya", "Deepika", "Chandran")
data = data.frame(Students, matrix1) 
data
```
The above data shows scores of different students in different subjects. The column names are automatically generated by R, however, the column names can be added as below. 
```{r}
headers=c("Students", "Geonomics", "Proteomics", "Microbiology", "Biostatistics")
colnames(data)=headers #add column names 
data
```

A rowwise addition can be performed on a data frame to find the total scores for each student in the four units
```{r}
## Add a new column with total marks obtained 
data$total_marks=rowSums(data[, c(2, 3, 4, 5)]) #add from second to fifth column 
data
```

Find the average score for each student.
```{r}
data$average_marks=rowMeans(data[, c(2, 3, 4, 5)])
data
```

















