# Data Types and Data Structures
## Data types 
<-- add more explanations to each data types --> 

- Numeric
```{r}
a=23.5
class(a) #check the data type of a 
```

- Integers
```{r}
a=23L #add L to show it is an integer
class(a)
```

- Characters
```{r}
a="DNA"
class(a)
```

- Logical 
```{r}
a=TRUE #logical data types is either TRUE or FALSE only
class(a)
```

- Complex, 
```{r}
a=2+3i # Complex data types have 'i' at the end of each number
class(a)
```

- Raw
```{r}
a=charToRaw("DNA")
print(a)
class(a)

# convert back to character 
b=rawToChar(a)
class(b)
```

## Data Structures 
This is the organization of data into one or multiple data values in specific structures. Different types of data structures in R include;
- Vector
- Matrix
- Data frame

### Vector
A vector is a single entity consisting of a collection of things. Here is an example of a vector.
```{r}
marks = c(23, 67, 98, 34, 98, 21)
print(marks)
```

The `class` function is utilized to determine the data types present within vector data values.
```{r}
class(marks)
```
The vector "marks" consist of only numeric values 

`is.vector` function is used to check if the variable is a vector 
```{r}
is.vector(marks)
```

unlike matrix and data frame, vector has no dimension 
```{r}
dim(marks)
```

`length()` function is used to count number of elements in vectors
```{r}
length(marks)
```

Index is the position of an element in a vector, in R it starts at index 1 - lets say we find the third element by index 3
```{r}
marks[3]
```
 value "98" is at index 3, or the third in the vector. 
 
 