# Loops
A loop is a sequence of instructions that are executed iterative until a condition is met. They are two types of loops in R, they are;
- for loops
- while loops 

## For loops 
This is a type of a loop that iterate over a sequence of elements, such as a vector, list, or sequence of numbers, executing a block of code for each element in the sequence. The loop iterates over each element of the sequence, one at a time, and executes the specified code block for each iteration.
Here is the syntax of a for loop 
```
for (variable in sequence) {
  # Code block to be executed for each iteration
}
```
When a condition is met the execution/iteration stops 
Below is an example of R loop over a range of numbers 1 to 10
```{r}
for (n in 1:10)
{
  print(n*5)
}
```

The loop can also iterate over a vector and perform the preset condition. For instance in the code below, the loops through a vector and prints a square of each element in the vector.
```{r}
num=c(1,2,3,8,10)
for (n in 1:length(num))
{
  print(num[n]^2)
}
```

Breaking of a loop - a program stops once something happens for instance the below loops executes by incrementing n and breaks when n is greater than 7
```{r}
#Stop the loop if the n is greater than 7
for (n in num)
{
  if (n > 7){
    break
  }
  
  print(n)
  
}
```

## While loop
A while loop is a control flow statement in programming that repeatedly executes a block of code as long as a specified condition is true. The loop continues to execute until the condition becomes false.

The below program executes as long as n is less than or equal to m. 

```{r}
m=4
n=-9
while(n<=m)
{
  print(n)
  n=n+1
}
```







