# Loops

<script src="js/dcl.js"></script>
<link rel="stylesheet" href="https://maxcdn.bootstrapcdn.com/bootstrap/3.4.1/css/bootstrap.min.css">
<script src="https://maxcdn.bootstrapcdn.com/bootstrap/3.4.1/js/bootstrap.min.js"></script>

<script src="js/dcl.js"></script>
```{r include=FALSE}
tutorial::go_interactive(greedy=TRUE)
knitr::opts_chunk$set(echo = TRUE,error=TRUE)
```


A loop is a sequence of instructions that are executed iterative until a condition is met. They are two types of loops in R, they are;
- for loops
- while loops 

## For loops 
This is a type of a loop that iterate over a sequence of elements, such as a vector, list, or sequence of numbers, executing a block of code for each element in the sequence. The loop iterates over each element of the sequence, one at a time, and executes the specified code block for each iteration.


Here is the syntax of a for loop;
```
for (variable in sequence) {
  # Code block to be executed for each iteration
}
```

When a condition is met the execution/iteration stops. The curly brackets `{`, `}` in the for-loop enclosed the code block to be be executed alliteratively to meet the said condition. 

Below is an example of R loop over a range of numbers 1 to 10
```{r}
for (n in 1:10)
{
  print(n*5)
}
```

Lets break down the above code;

1. `for (n in 1:10)`

This line initializes a for loop that iterates over the numbers from 1 to 10. The 1:10 sequence generates the numbers 1, 2, 3, ..., 10.
The variable n takes on each value in this sequence in turn, starting from 1 and ending at 10

2. `{...}`

This is the loop body that contains the statement/code that will be executed alliteratively until `n` is 10. 

3. `print(n*5)`

- Inside the loop body, the print() function is used to output the result of the expression n*5 to the console.
- n*5 multiplies the current value of n by 5

In summary, the for loop above iterates through the numbers 1 to 10 and prints the respective product with 5 to the console. 

<span style="color: green;">**Activity**</span>

Using a for-loop, print each number from 10 to 20 alliteratively

```{r ex="activity-3a", type="sample-code"}
# loop from 10 to 20
# CODE HERE
```

```{r ex="activity-3a", type="solution"}
# loop from 10 to 20
for (i in 10:20)
{
  print(i)
}
```

**Loops in vectors**

Not only can for loops go through a set of numeric values but also they can loop through a vector of characters. For instance;
```{r}
# This will loop through a set of fruits
fruits <- c("apples", "bananas", "oranges")
for (fruit in fruits) {
  print(paste("I like", fruit))
}
```

Lets break down the code below; 

1. `fruits <- c("apples", "bananas", "oranges")`

This is vector of fruits that will be efficiently iterated in the for loop. 

2. `for (fruit in fruits)`

This line initializes a loop that iterates through the vector `fruits`. 

3. `{...}` the loop body. 

It contains code statement that will be executed. 

4. `print(paste("I like", fruit))`

Each fruit in the vector is printed out im the console along with the statement "I like". 


<span style="color: green;">**Activity**</span>

Create a vector variable `cars` and store the following values; `"Mazda", "Toyota", "Ford", "Ferrari","Isuzu"`. Iterate through `cars` and print out "I bought a ____". For the blank input a car model. 

```{r ex="activity-3b", type="sample-code"}
# Create a vector cars
# CODE HERE 

# Loop through the vectors and print "I bought a ____" 
# CODE HERE 
```

```{r ex="activity-3b", type="solution"}
# Create a vector cars
cars = c("Mazda", "Toyota", "Ford", "Ferrari","Isuzu")

# Loop through the vectors and print "I bought a ____" 
for (car in cars)
{
  print(paste("I bought a", car))
}
```

**Nested loops**

A for loop can run within another for loop, this is referred to as a "Nested for loop". The code below shows how nested for loops work
```{r}
for (i in 1:3) {
  for (j in 1:2) {
    print(paste("i:", i, ", j:", j))
  }
}
```

The code `for (j in 1:2) { ...}` is nested within `for (i in 1:3) {...}` and the former is executed in each iteration of the later. 


These nested for loops are very useful when working with multidimensional data structures like matrices. Like matrix by matrix multiplication;

- Example 1
```{r}
# Matrix multiplication
matrix1 <- matrix(1:4, nrow = 2)
matrix1
matrix2 <- matrix(5:8, nrow = 2)
matrix2

# Matrices are multiplied row(of matrix1) by column(of matrix2)
result <- matrix(0, nrow = nrow(matrix1), ncol = ncol(matrix2))
for (i in 1:nrow(matrix1)) {
  for (j in 1:ncol(matrix2)) {
    for (k in 1:ncol(matrix1)) {
      result[i, j] <- result[i, j] + matrix1[i, k] * matrix2[k, j]
    }
  }
}
print(result)
```

- Example 2
```{r}
# Define two matrices
matrix1 <- matrix(1:9, nrow = 3)
matrix1
matrix2 <- matrix(9:1, nrow = 3)
matrix2

# Initialize result matrix
result <- matrix(0, nrow = nrow(matrix1), ncol = ncol(matrix2))

# Perform matrix multiplication using nested for loops
for (i in 1:nrow(matrix1)) {
  for (j in 1:ncol(matrix2)) {
    for (k in 1:ncol(matrix1)) {
      result[i, j] <- result[i, j] + matrix1[i, k] * matrix2[k, j]
    }
  }
}

result
```

The loop can also iterate over a vector and perform the preset condition. For instance in the code below, the loops through a vector and prints a square of each element in the vector.
```{r}
num=c(1,2,3,8,10)
for (n in 1:length(num))
{
  print(num[n]^2)
}
```

<span style="color: green;">**Activity**</span>

- Create a sequence of numbers from 1 to 3 and store it in a variable `x`.
 
- Create a sequence of numbers from 1 to 5 and store it in a variable `y`.

- Write a nested loop:

- The outer loop should iterate through each value in `x`.

- The inner loop should iterate through each value in `y`.

- Inside the inner loop, print the current values of `x` and `y` in the format: `"x = <value>, y = <value>"`.

```{r ex="activity-3c", type="sample-code"}
# Create a sequence of numbers from 1 to 3 and store it in a variable `x`.
# CODE HERE

# Create a sequence of numbers from 1 to 5 and store it in a variable `y`.
# CODE HERE 

# Write a nested loop
# The outer loop should iterate through each value in `x`.
# The inner loop should iterate through each value in `y`.
# - Inside the inner loop, print the current values of `x` and `y` 
# in the format: `"x = <value>, y = <value>"`.
# CODE HERE 
```

```{r ex="activity-3c", type="solution"}
# Create a sequence of numbers from 1 to 3 and store it in a variable `x`.
x = seq(1:3)

# Create a sequence of numbers from 1 to 5 and store it in a variable `y`.
y = seq(1:5)

# Write a nested loop
# The outer loop should iterate through each value in `x`.
# The inner loop should iterate through each value in `y`.
# - Inside the inner loop, print the current values of `x` and `y` 
# in the format: `"x = <value>, y = <value>"`.

# The outer loop should iterate through each value in `x`.
for (xi in x){
  # The inner loop should iterate through each value in `y`.
  for (yj in y){
    # Inside the inner loop, print the current values of `x` and `y` 
    print(paste("x =", xi, "y =", yj))
  }
}
```


**Breaking of a loop**

A program stops once something happens for instance the below program executes by looping through a vector `num` and breaks when `n` is greater than 7. 
```{r}
num=c(1,2,3,8,10)

#Stop the loop if the n is greater than 7
for (n in num)
{
  if (n > 7){
    break
  }
  
  print(n)

}
```

<span style="color: green;">**Activity**</span>

- Create a vector `fruits` which have the elements; `"mango", "avocado" , "pineapple" , "peach", "banana", "orange"`. 

- Loop through the `fruits` , print each element

- Break the loop once you print 'banana'. 

```{r ex="activity-3d", type="sample-code"}
# Create a vector fruits 
# CODE HERE

# Loop through the `fruits` , print each element
# Break the loop once you print 'banana'.
# CODE HERE 
```

```{r ex="activity-3d", type="solution"}
# Create a vector fruits 
fruits = c("mango", "avocado" , "pineapple" , "peach", "banana", "orange")

# Loop through the `fruits` , print each element
# Break the loop once you print 'banana'.

# Loop through the `fruits` , print each element
for (fruit in fruits)
{
  print(fruit)
  # Break the loop once you print 'banana'.
  if(fruit=="banana")
  {
    break
  }
}
```


## While loop
A while loop is a control flow statement in programming that repeatedly executes a block of code as long as a specified condition is true.
After executing the code block, the program returns at the beginning the condition, if its is TRUE the code block is executed again. 

The loop continues to execute until the condition becomes false.
```
while (condition) {
  # Code block to be executed
}
```

- Example 1

The below program executes as long as `n` is less than or equal to `m`. 

```{r}
m=4
n=-9
while(n<=m)
{
  print(n)
  n=n+1
}
```

The condition of the above code is that the program will executes as long as `n` is less than or equal to `m`. The value of `n` is incremented by 1 and printed out until as long its less than or equal to `m`. 

- Example 2

```{r}
i = 10
while (i >= 5) {
  print(i)
  i = i - 1
}
```

This while loop prints the values of `i` from 10 to 5. It starts with ``i` equal to 10 and continues iterating as long as i is more than or equal to 5. Inside the loop, each value of `i` is printed, and `i` is reduced by 1 in each iteration.

<span style="color: green;">**Activity**</span>

Learn how to use a `while` loop to repeatedly perform an operation as long as a condition is true;

- Create a variable `b` and assign it the value 5.
- Use a while loop to repeatedly add 1 to `b` and print the result.
- The loop should continue as long as `b` is less than or equal to 15.

```{r ex="activity-3e", type="sample-code"}
# Create a variable `b` and assign it the value 5
# CODE HERE

# Use a while loop to repeatedly add 1 to `b` and print the result.
# The loop should continue as long as `b` is less than or equal to 15
# CODE HERE 
```

```{r ex="activity-3e", type="solution"}
# Create a variable `b` and assign it the value 5
b = 5

# Use a while loop to repeatedly add 1 to `b` and print the result.
# The loop should continue as long as `b` is less than or equal to 15

while (b<=15)
{
  print(b)
  b = b+1
}
```

**Infinite Loop**: If the condition in a while loop always evaluates to TRUE, the loop becomes an infinite loop, which continuously executes the code block without stopping. Infinite loops should be avoided, as they can cause the program to hang or become unresponsive.

There are different use cases in R while loops. These are a few examples of use cases;

1. Calculating factorial
```{r}
# Function to calculate factorial
factorial = function(n) {
  result = 1
  i = 1
  while (i <= n) {
    result = result * i
    i = i + 1
  }
  return(result)
}

# Calculate factorial of 5
factorial_5 = factorial(5)
print(factorial_5)  # Output: 120
```

2. Use input validation
```{r}
# Prompt user to enter a number until a valid numeric value is provided
num = NA
while (is.na(num)) {
  input = 3
  num = as.numeric(input)
  if (is.na(num)) {
    cat("Invalid input. Please enter a numeric value.\n")
  }
}
print(paste("You entered:", num))
```

3. Generating a fibonacci series
```{r}
# Generate Fibonacci series up to n terms
fibonacci = function(n) {
  a = 0
  b = 1
  i = 1
  while (i <= n) {
    cat(a, " ")
    next_term = a + b
    a = b
    b = next_term
    i = i + 1
  }
  cat("\n")
}

# Generate Fibonacci series with 10 terms
fibonacci(10)  # Output: 0  1  1  2  3  5  8  13  21  34
```


## CODE CHALLENGE
Solve the following equations using `for` and `while` loops;

1. A vector `v` contains `34, 86, 23, 65, 98`. Use a `for` loop to print each `v` element. 
2. Create a matrix with the first column containing numbers from 1 to 3 and b containing 4 to 6. Use a `for` loop to print each row of the data frame. 
3. Vector `x` has elements, `1, 2, 3` while vector `y` consist of elements, `3, 0, -3`. Use nested `for` loops to print all combinations of the elements in `x` and `y`. 
4. Create a sequence, `m`, from 1 to 10. Use a `for` loop to print each element, but break the loop if the element is equal to 5.
5. A variable `b` initially has a value of 1. Double the value of `b` alliteratively and print it out to the console. The printing should stop once `b` is greater than or equal to 30. Use a `while` loop for this. 
6. A vector `names` contains elements, `"Vipin", "Patel", "Pradip", "Salman", "Tom"`. Iterate through the vector and print `"Hello"` alongside each name to print greet the participant. 

**Hint**: *Use a for loop*

7. Create a matrix `m` filled with values from 1 to 12 that has 4 rows and 3 columns. Use a for loop to iterate through each row and find its sum.
8. A variable `n` has a value 3. Add 3 to `n` alliteratively and print out the result as long as `n` is equal to or less than 25. Use a `while` loop to solve this problem. 
9. A vector `v` contains values from 1 to 7. Using a `for` loop, iterate through each element the use a `while` loop to subtract `1` from each element until it reaches 0. Print out the result(It should print all zeros). 

**Hint**: *Nest a ``while` loop in a `for` loop*.

10. Create a vector, `animals` containing `"cat", "lion", "dog", "sheep"`. Use a for loop to print out each element alongside the number of characters the element contains. Use `nchar` to count number of characters in character data types 