# Dpyr library

<script src="js/dcl.js"></script>
<link rel="stylesheet" href="https://maxcdn.bootstrapcdn.com/bootstrap/3.4.1/css/bootstrap.min.css">
<script src="https://maxcdn.bootstrapcdn.com/bootstrap/3.4.1/js/bootstrap.min.js"></script>

<script src="js/dcl.js"></script>
```{r include=FALSE}
tutorial::go_interactive(greedy=TRUE)
knitr::opts_chunk$set(echo = TRUE,error=TRUE)
```

`readxl` library has been used to read excel table into R and manipulate the data using R basic functions, now `dplyr` library will be used. This, `dplyr` library is an advanced package that has functions used to manipulate data. The functions are referred to as dplyr verbs. This are some of the examples of dplyr verbs;

- select
- filter
- arrange 
- mutate
- group by 
- summarise

To install `dplyr`, run the below command on the console.
```
install.packages("dplyr")
```
Ensure your computer is connected to the internet when installing!

To invoke the libraries to the program run;
```{r echo=TRUE, error=FALSE, warning=FALSE, message=FALSE}
library(dplyr)
library(readxl)
```

Before working on the data, it need to imported first. The `readxl` library will be used. The data will then be manipulated by `dplyr`.


Apart from data retrieved from excel tables, there are some inbuilt freely available datasets in R that a student can use for practice. `data()` function can be used to view some of the freely available datasets in R. 
```{r}
data()
```

This will show all data sets including both the inbuilt and the previously imported ones. Alternatively the code below will show only the inbuilt R data sets.
```{r}
ls("package:datasets")
```

Some libraries come with their own inbuilt data sets which can be explored by the command;
```{r}
data(package=.packages(all.available = TRUE))
```

Good example is the famous iris that will be used in this course. It is always used by beginners in data science. The data set contains measurements of various characteristics of iris flowers. These characteristics include sepal length, sepal width, petal length, and petal width. There are three species of iris flowers in the data set: setosa, versicolor, and virginica. 
The data will be invoked to R before assessment and wrangling;

Lets load the iris data and explore the first few records. 
```{r}
data("iris")
head(iris)
```

Calculating the summary statistics
```{r}
summary(iris)
```

Lets write this data set locally to a file. 
```{r}
write.table(iris, "data/iris_dataset.txt", append = FALSE)
file.show("data/iris_dataset.txt")
```

or to remove quotes next to values 
```{r}
write.table(iris, "data/iris_dataset.txt", append = FALSE, quote = FALSE)
file.show("data/iris_dataset.txt")
```

or to remove quotes and row names
```{r}
write.table(iris, "data/iris_dataset.txt", append = FALSE, quote=FALSE, row.names = FALSE)
file.show("data/iris_dataset.txt")
```

## Dplyr Verbs
### select

This dplyr verb is used when selecting or dropping specific columns. In this lesson we will find the iris column names and select two of them using `select`. 
```{r}
# Find the column names 
colnames(iris)
```

Remember the data frame to work on need to be specified in the arguments such that 

```
selected_data = select(data_frame, col1, col2, col3)
```

Therefore, we will select the columns; Species, Petal length and petal width. 

```{r}
selected_iris_data = select(iris,
                       Petal.Length, Petal.Width, Species)

# view the first few rows of the selected data
head(selected_iris_data)
```
The three selected columns are displayed in the data frame above. 

Specific columns can be dropped by putting `-` before the column name as
```
# Drop specified columns 
remaining_data = select(data_frame, -col1_to_drop, -col2_to_drop)
```
In this lesson, we will drop petal length, petal width and Species columns;
```{r}
remaining_iris_data = select(iris,
                       -Petal.Length, -Petal.Width, -Species)

# view the first few rows of the selected data
head(remaining_iris_data)
```

<span style="color: green;">**Activity**</span>

You will be required to use the `car_sales` data set from https://raw.githubusercontent.com/insaid2018/Term-1/master/Data/Projects/car_sales.csv.Read the data using `read.csv` and select the `car, price, body, mileage, engV, engType, year, model`. Save the data frame from the selected columns as `selected_cars_df`. Show the first few rows of the `selected_cars_df`. 

```{r ex="activity-5a", type="sample-code"}
# Load the dplyr library
# CODE HERE

# Read the data 
# CODE HERE

# select the required columns 
# CODE HERE

# Show the first few rows of the data frame
# CODE HERE
```

```{r ex="activity-5a", type="solution"}
# Load the dplyr library
library(dplyr)

# Read the data 
data_link = "https://raw.githubusercontent.com/insaid2018/Term-1/master/Data/Projects/car_sales.csv"
car_sales = read.csv(data_link)

# select the required columns 
selected_cars_df = select(car_sales, 
                          car, price, body, mileage, engV, engType, year, model)

# Show the first few rows of the data frame 
head(selected_cars_df)
```

### filter
Is a verb/function from `dplyr` used to filter records in a data frame based on a specific condition. It allows the analyst to retrieve the records he/she is interested in and work easier with the subset. 
With `filter()`, the data frame and the condion are passed as a arguments;
```
# Filtering rows where a certain column meets a condition
filtered_data = filter(data_frame, 
column_name > 5 # This is the condition)
```

Lets select the species 'setosa' from the iris data set 
```{r}
setosa_iris = filter(iris, # the data frame
                     Species == "setosa" # the condition
                     )
# First few records of setosa data
head(setosa_iris)
```

Records with sepal width of more than 3.0 can be filtered. Here is how we achieve such a subset
```{r}
wide_sepal_iris = filter(iris, #the data frame
                         Sepal.Width>3.0 # the condition
                         )

head(wide_sepal_iris)
```

<span style="color: green;">**Activity**</span>

With the `car_sales` data set that you used above, use `filter()` function to get the cars that were sold from the year 2008 to present and name them `latest_car_sales`. Count the number of observations made and show the first few rows. 

```{r ex="activity-5b", type="sample-code"}
# Load the dplyr library
# CODE HERE 

# Read the data 
# CODE HERE

# Filter to find cars sold from 2008
# CODE HERE 

# Count the observations made. Use nrows function
# CODE HERE 

# Show the first few rows
# CODE HERE 
```

```{r ex="activity-5b", type="sample-code"}
# Load the dplyr library
library(dplyr)

# Read the data 
data_link = "https://raw.githubusercontent.com/insaid2018/Term-1/master/Data/Projects/car_sales.csv"
car_sales = read.csv(data_link)

# Filter to find cars sold from 2008
latest_car_sales = filter(car_sales, 
                          year>=2008)

# Count the observations made. Use nrow function
nrow(latest_car_sales)

# Show the first few rows
head(latest_car_sales)
```

### arrange 
This is `dplyr` verb/function used for sorting rows by rearranging in a specific order. here is how to use `arrange()` function;
```
arranged_data = arrange(data_frame, column_name)
```
This allows the analyst to arrange the data in a default ascending order. To arrange in a descending order a `desc()` function is added as;
```
# Note the additional desc function
arranged_data = arrange(data_frame, desc(column_name))
```

Now lets order the iris data in an ascending order based on Petal length and view the first 6 records with the shortest petal. 
```{r}
# Sort the data 
by_petal_length = arrange(iris, # data frame 
                          Petal.Length # order by column
                          )

# View the data 
head(by_petal_length)
```

Lets repeat the same process but now we order the data in a descending order.
```{r}
# Sort the data 
by_petal_length = arrange(iris, # data frame 
                          desc(Petal.Length) # order by column
                          )

# View the data 
head(by_petal_length)
```

<span style="color: green;">**Activity**</span>

Arrange the columns in the `car_sales` data set according to `mileage` in descending order. Show the last few rows

```{r ex="activity-5c", type="sample-code"}
# Load the dplyr library
# CODE HERE 

# Read the data into a data frame
# CODE HERE 

# Order according to mileage in descending order
# CODE HERE 

# Show the last few rows of the data set
# CODE HERE
```

```{r ex="activity-5c", type="solution"}
# Load the dplyr library
library(dplyr)

# Read the data into a data frame
data_link = "https://raw.githubusercontent.com/insaid2018/Term-1/master/Data/Projects/car_sales.csv"
car_sales = read.csv(data_link)

# Order according to mileage in descending order
mileage_order = arrange(car_sales, 
                        desc(mileage))

# Show the last few rows of the data set
tail(mileage_order)
```


### mutate
`mutate()` is a dplyr verb used to modifying the existing variables or creating new variables in a data set. 
In this case we can calculate the log off Sepal length in the iris data 
```{r}
# modify Sepal.Length
new_iris = mutate(iris, Sepal.Length=log(Sepal.Length))

head(new_iris)
```

Additionally, we can create an entirely new variable by `mutate()`. In this case we will find the ratio between petal length and petal width. The new variable will be called "Petal.Length.Width.Ratio"
```{r}
new_iris = mutate(iris,
                  Petal.Length.Width.Ratio = Petal.Length/Petal.Width)

head(new_iris)
```

The "Petal.Length.Width.Ratio" is found by dividing the Petal.Length and the Petal.Width variables. 

<span style="color: green;">**Activity**</span>

Using the `car_sales` data set, create a new column, `"distance_covered_km"`, calculated from the `mileage`. Just multiply `mileage` with `1.609`. Show the first few rows of the mutated data frame. 

```{r ex="activity-5d", type="sample-code"}
# Load the dplyr library
# CODE HERE 

# Read the data set 
# CODE HERE 

# Create new column "distance_covered_km
# CODE HERE 

# Show first few rows 
# CODE HERE 
```

```{r ex="activity-5d", type="solution"}
# Load the dplyr library
library(dplyr)

# Read the data set 
data_link = "https://raw.githubusercontent.com/insaid2018/Term-1/master/Data/Projects/car_sales.csv"
car_sales = read.csv(data_link)

# Create new column "distance_covered_km
mutated_car_sales = mutate(car_sales,
                           distance_covered_km = mileage * 1.609)

# Show first few rows 
head(mutated_car_sales)
```

### group_by 
The `group_by()` is a function used to group records in a data frame by one or more variables. It allows the analyst to create a group based on a certain criteria. 

Lets group the iris data based on the Species variable;
```{r}
iris_groups = group_by(iris, Species)
head(iris_groups)
```

This groupings allow the analyst to retrieve insights at more base level and uncover more insights that could not have been possible when analyzing the entire data set

<span style="color: green;">**Activity**</span>

Use the `car_sales` data set provided before to work on this activity. Load the data and group the sales by `model` to get the sum of every quantitative feature/variable. Name the resultant data frame, `car_sales_by_model`. Display the resultant data frame. 

```{r ex="actvity-5e", type="sample-code"}
# Load the dplyr library
# CODE HERE 

# Load the car sales data 
# CODE HERE 

# Group the sales by model
# CODE HERE 

# Print out the dataframe 
# CODE HERE 
```

```{r ex="actvity-5e", type="solution"}
# Load the dplyr library
library(dplyr)

# Load the car sales data 
data_link = "https://raw.githubusercontent.com/insaid2018/Term-1/master/Data/Projects/car_sales.csv"
car_sales = read.csv(data_link)

# Group the sales by model
car_sales_by_model = group_by(car_sales, model)

# Print out the dataframe 
car_sales_by_model
```


### Summarise
To calculate summary statistics such as average, median and maximum the `summarise()` is used. This function collapses mutilple rows into a summary row. For instance calculating the mean Petal width;
```{r}
summarise(iris, 
          mean_petal_width=mean(Petal.Width))
```

To find the mean petal width for each iris species;
- the iris data will be grouped by species
- a mean value for each group will be calculated
```{r}
grouped_iris = group_by(iris, Species)
mean_petal_widths = summarise(grouped_iris, mean_value=mean(Petal.Width))
mean_petal_widths
```

<span style="color: green;">**Activity**</span>

You will be required to use the `car_sales` data set once again. Calculate the descriptive statistics using `summarise()` command. 

```{r ex="activity-5f", type="sample-code"}
# Load the dplyr library
# CODE HERE 

# Load the car sales data 
# CODE HERE 

# Calculate the summary statistics
# CODE HERE
```

```{r ex="activity-5f", type="solution"}
# Load the dplyr library
library(dplyr)

# Load the car sales data 
data_link = "https://raw.githubusercontent.com/insaid2018/Term-1/master/Data/Projects/car_sales.csv"
car_sales = read.csv(data_link)

# Calculate the summary statistics
summarise(car_sales)
```

## Chaining
Chaining is the process of combining several operations together using the `%>%` or forward pipe operator. The chained workflow succeeds each other until the whole process is done. To understand chaining, the mtcars(Motor Trend cars) data set will be used. Mtcars is also a well-known data set containing several attributes of 32 different cars from 1974. Here's a brief explanation of the variables in the mtcars data set:

1. mpg: Miles per gallon (fuel efficiency).
2. cyl: Number of cylinders.
3. disp: Displacement (cubic inches).
4. hp: Horsepower.
5. drat: Rear axle ratio.
6. wt: Weight (in 1000 lbs).
7. qsec: Quarter mile time (in seconds).
8. vs: Engine type (0 = V-shaped, 1 = Straight).
9. am: Transmission type (0 = Automatic, 1 = Manual).
10. gear: Number of forward gears.
11. carb: Number of carburetors.

Lets start by loading the data into the program and view its first few records;
```{r}
data(mtcars)
head(mtcars)
```

Lets `select` 6 most important columns in this analysis
```{r}
cars1 = mtcars %>% select(mpg, cyl, disp, hp, qsec, am)

head(cars1)
```

Lets now `filter` to find vehicles with an automatic transmission type. The `filter` verb will be chained to `select` verb with `%>%`. 
```{r}
cars2 = mtcars %>%select(mpg, cyl, disp, hp, qsec, am) %>%
  filter(am==0)
head(cars2)
```

All these vehicles are of automatic transmission type, lets rank them according to the horsepower in descending order. 
```{r}
cars3= mtcars %>%select(mpg, cyl, disp, hp, qsec, am, wt) %>%
  filter(am==0) %>%
  arrange(desc(hp))

head(cars3)
```

A new column of weight in 1000kgs (wt_1000kgs) can be created by diving weight in 1000lbs by 2.20462. `mutate` verb will be chained also.
```{r}
cars4= mtcars %>%select(mpg, cyl, disp, hp, qsec, am, wt) %>%
  filter(am==0) %>%
  arrange(desc(hp)) %>%
  mutate(wt_1000kgs=wt/2.20462)

head(cars4)
```

The above process has explained how chained works in `dplyr`. Many functions/processed can be chained together to manipulate data to the desired output. The next section will apply chaining to biology and be used to answer a few questions that will cement your understanding in R as a biologist. 

## Application of Dplyr on Microbiology
Genetics composition of an organism can be analyzed using `dplyr`, in this course unit it will be used to analyze the human genetics data. 

- Read the data set from a file into a data frame
```{r}
hg = read_excel("data/Human_genome.xlsx")
class(hg)
```

review the first six records of the data set 
```{r}
head(hg)
```
 
 the last six records of the data set
```{r}
tail(hg)
```
 
`sample_n()` command is used to randomly sample fixed number of rows from the data frame. Lets sample five rows randomly
```{r}
hg %>% sample_n(5)
```
 
`sample_fra()` command can also be used to extract a percent of rows randomly. Lets samople 25% of the rows 
```{r}
hg %>% sample_frac(0.25, replace = FALSE)
```

The `replace` argument allows repetition of rows when sampling when set `TRUE` and vice versa. The essence of `dplyr` in microbiology data can be explained better by questions.

**Q1. Which chromosomes has maximum size**

The functions used here are; select, arrange, slice 

1. The required columns will be selected i.e Chromosomes and Base_Pairs(size)
2. The data will ordered according to Base_Pairs in descending order by the `arrange` command.  
3. the data will be "sliced" to select the first row 

All these processes will be chained to run progressively from process 1 to 3. 

All the above steps should be piped
```{r}
hg1 = hg %>% #transfer the data to a new variable
  select(Chromosome, Base_Pairs) %>% # select the required columns 
           arrange(desc(Base_Pairs)) %>% #order in descending order
           slice(1)

hg1
```

**Q2. List of Chromosomes with more than 500 pseudogenes and more than 1000 protein coding genes**

The human genome data frame will be filtered to find the Chromosomes with more than 500 pseudogenes and more than 1000 protein coding genes.
`&` condition is used in the filter 

```{r}
hg2=hg%>%
  filter(Pseudogenes>500 & Protein_Coding_genes>1000)
hg2
```

**Q3. List of Chromosomes with more tha 500 pseudogenes or more than 1000 protein coding genes and miRNA more than 100**

a frame will be filtered to find the Chromosomes with more than 500 pseudogenes 
The human genome dator more than 1000 protein coding genes and mtRNA more than 100.

The `&` and `|` conditions will be used in place of "and" and "or" respectively.
```{r}
hg2=hg %>% 
  filter((Pseudogenes>500|Protein_Coding_genes>1000) & miRNA>100)
hg2
```

**Q4. Which 10 Chromosomes have the highest protein gene density (Genes per megabase)**

- Mutate -> is a command to create a new column 

Lets create a column of genes per megabase where a megabase is 1000000 base pairs
```{r}
hg3=hg %>%
  mutate(Genes_per_Mb=round((Protein_Coding_genes/(Base_Pairs/100000)),
                            digit=2))
hg3
```

then order the data according to genes per megabse column (`Genes_per_mb`) in descending order and select the first 10 rows using `arrange` and `select` commands. 
```{r}
hg4=hg3 %>%
  arrange(desc(Genes_per_Mb)) %>% # order in descending order 
  select(Chromosome, Genes_per_Mb) %>% # select the required columns 
  slice(1:10) # select the top 10

hg4
```

**Q5. Calculate the mean of gene density for autosomes, sex chromosomes and mitochondrial genes**

```{r}
autosomes=rep("A", 22) #create a vector with 22 As
print(autosomes)
chr_class=c(autosomes, "S", "S", "M")
chr_class
hg5= hg3 %>%
  mutate(Type=chr_class)

hg5 # confirm the new column "Type" is added
```

Group by chromosomes and calculate the mean density 
```{r}
hg6=hg5 %>% 
  group_by(Type) %>%
  summarise(gene_density=mean(Genes_per_Mb))
hg6
```

## CODE CHALLENGE
Palmer Archipelago(Antarctica) Penguin data - https://www.kaggle.com/datasets/parulpandey/palmer-archipelago-antarctica-penguin-data

1. `ToothGrowth` is one of the freely available data sets in base R. Go on and;
    a. Load the `ToothGrowth` data set. 
    b. Display the first 5 records.
    c. Use the function `write.table` to write the data set to a local file. 
2. You will use the Palmer Archipelago(Antarctica) Penguin data set from kaggle to practice various data manipulation techniques using `dplyr` verbs in R. Ensure You load the data as `penguin_data` and necessary packages before working on the following questions below;
    a. Using `select()` verb, create a new data set that contains only the columns `Species, Island, Flipper.Length..mm.` and name the dataframe `selected_features`. Show the first 10 observations of `selected_features`.
    b. Filter the Penguin data set to retain only the "Adelie" species that live in "Torgersen" island. Use `filter()` function. Name the resultant data set `Adelie_data` and show its first 7 records. 
    c. Arrange the Penguin data set according to `Flipper.Length..mm.` in descending order. Use `arrange()` verb. Name the resultant data frame, `ordered_penguin_data` and show the first five rows.
    d. Using the `mutate()` function, add a new column called `Flipper.Length..mm.` derived from `Flipper.Length..mm.` by 10. Name the data set, `modified_penguin_data`. and show the first rows
    e. Group the data set according to `island`. Name the resultant data set, `penguin_data_by_island` and print the data(first 10 rows) to the console.
    f. For each species and island, calculate the mean `Flipper.Length..mm.` and the observations count. 
3. In this activity, you will chain at least four `dplyr` verbs to achieve the desired outcome. 
    - Select the columns, `island, body_mass_g, species, bill_length_mm`
    - Filter the selected data set to include penguins that have a bill length greater than 40mm
    - Create a new column/variable that will convert `bill_length_mm` to `bill_length_cm` i.e divide `bill_length_mm` by 10 to come up with `bill_length_cm`. 
    - Order the resulting data set by `body_mass_g` in descending order. 