# Video 5

# ADVANCED LEVEL 

<--EXPLAIN THE PROCESS OF dplyr INSTALLATION USING RSTUDIO -->

```{r}
#uncomment the below line to install dplyr
#install.packages("dplyr")
```


Introduce dplyr and dplyr verbs 
- select
- filter
- arrange 
- mutate
- group by 
- summarise

Set your working directory

Load the required libraries 
```{r}
library(dplyr)
library(readxl)
```

Now we can access commands both in  `readxl` and `dplyr`

```{r}
hg = read_excel("data/Human_genome.xlsx")
View(hg)
```

What are the inbuilt datasets in R
```{r}
data()
```

or 
```{r}
ls("package:datasets")
```


data available in all packages 
```{r}
data(package=.packages(all.available = TRUE))
```

Example - working with iris dataset
```{r}
data("iris")
View(iris)
```

summary statistics 
```{r}
summary(iris)
```

writing to a file 
```{r}
write.table(iris, "data/iris_dataset.txt", append = FALSE)
file.show("data/iris_dataset.txt")
```

or to remove quotes next to values 
```{r}
write.table(iris, "data/iris_dataset.txt", append = FALSE, quote = FALSE)
file.show("data/iris_dataset.txt")
```

or to remove quotes and row names <-- remember to add screenshots--> 
```{r}
write.table(iris, "data/iris_dataset.txt", append = FALSE, quote=FALSE, row.names = FALSE)
file.show("data/iris_dataset.txt")
```
