# GGPLOT

<script src="js/dcl.js"></script>
<link rel="stylesheet" href="https://maxcdn.bootstrapcdn.com/bootstrap/3.4.1/css/bootstrap.min.css">
<script src="https://maxcdn.bootstrapcdn.com/bootstrap/3.4.1/js/bootstrap.min.js"></script>

<script src="js/dcl.js"></script>
```{r include=FALSE}
tutorial::go_interactive(greedy=TRUE)
knitr::opts_chunk$set(echo = TRUE,error=TRUE)
```

`ggplot2` is a powerful data visualization package in R that allows you to create a wide variety of graphs and plots. It is based on the grammar of graphics, a structured approach to constructing plots that emphasizes the mapping of data to visual attributes.

## Five key steps to ggplot 
There are 5 key steps in plotting in ggplot;
<p>
  1.**The Setup** - Read the data set, define x and y axis 
  
  2. **The Labels** - Title, X and Y axis labels 
  
  3. **The Theme** - Default, Black and White, colored etc. 
  
  4. **The Facets** - Individual Graphs for each group in data with exactly same range
  
  5. **The Layers or geoms** - The actual plot type - e.g Bar plot, Box plot, Violin plot etc. 
</p>

To install `ggplot`, run the below command in the console.  
```
install.packages("ggplot2")
```


## Prepare the data for plotting - Data Assessment
Invoke the libraries 
```{r echo=TRUE, error=FALSE, warning=FALSE, message=FALSE}
library(ggplot2)
library(data.table)
```

Read the data using the fread function 
```{r}
methylated_c = fread("data/R_WORKSHOP.csv", header=T, sep=",")
```

View the first five records of the data 
```{r}
head(methylated_c)
```

Look at the nature of the data 
```{r}
str(methylated_c)
```
In the data set, there are numeric variables a well as character columns. There are 15 records in the data set

Calculate the summary statistics 
```{r}
summary(methylated_c)
```

Do the data have null values?
```{r}
sum(is.na(methylated_c))
```

This data does not have any null values. 

```{r}
class(methylated_c)
```


## Plotting 
### Base set up
Before plotting the data, you need to;

- Define the graph skeleton. This is just an empty `ggplot()` function without additional arguments

- Define the aesthetics. It defines how data should be mapped to visual properties like position, color, shape, and size

- Define the labels. The labels make the visualizations informative and easy to understand for the user. 

```{r}
graph=ggplot(methylated_c, 
             aes(x=Chromosome, y=Count_in_millions))+ #graph aesthetics
  labs(x="Chromosomes", y="Cytosine count in millions", title="C distribution in A thaliana") #graph labels

print(graph)
```

- Define the type of graph - geoms. For this case, the graph is a bar graph. `geoms()` stands for 'geometric objects'. In `ggplot`, there are different types of `geoms` to define which graph is to be plotted. Some of the examples include; 

- `geom_point()` for scatter plot

- `geom_line()` for a line chart 

- `geom_density()` for a density plots

- `geom_histogram()` for histograms

- `geom_boxplot()` for box plots

- `geom_col()` for bar chart

- `geom_tile()` for heat maps

In this course we shall explore a few of them but the concepts applies to all of them. Now lets make our plot a bar chart. `geom()` function will be added to a variable `graph` defined above to plot a bar chart. 
```{r}
graph1 = graph + geom_col() #plot a bar chart
print(graph1)
```

If you can remember our data, `methylated_c` has a column/variable "Context". In this step, the variable "Context" will be introduced to make a stacked bar chart, therefore, the C distribution in Chromosomes based on "Context" will be plotted to add some depth to the data. 
```{r}
graph2=graph1+aes(fill=Context)
print(graph2)
```

To make it a grouped bar chart, a `position` argument is introduced. 

```{r}
graph3=graph+geom_col(position="dodge")+aes(fill=Context)
print(graph3)
```

Faceted graphs - these are related data plotted separately on the same image/diagram.

The `facet_grid()` function is added to the variable `graph3` to created faceted graphs based on the variable "Context". 
```{r}
graph4=graph3+facet_grid(~Context)
print(graph4)
```

GGPlot randomly select colors to paint the graph but the biologist/statistician can use predefined colors as below.
```{r}
graph5=graph4+scale_fill_manual(values=c("brown", "blue", "red"))
print(graph5)
```

- The background color can be modified to ones aesthetic taste. The `theme()` function along with relevant arguments to make the background color of your choice, in this its light blue. 
```{r}
graph5+theme_bw(base_size=9)+theme(panel.background = element_rect(fill="lightblue"))
```

All of the above steps can be combined to come up with a decent graph from ggplot library.
```{r}
my_title = expression(paste("Cystosine distribution in ", italic("Arabidopis thaliana")))#Define the title 

graph_final = ggplot(methylated_c, aes(x=Chromosome, y=Count_in_millions))+
  labs(x="Chromosomes", title = my_title) +
  facet_grid(~Context) +
  geom_col(position="dodge")+
  aes(fill=Context)+scale_fill_manual(values = c("brown", "blue", "red"))+
  theme_bw(base_size = 9)+theme(panel.background = element_rect(fill="lightblue"))

print(graph_final)
```
The `expression` is used to combine name italics and normal fonts - this helps biologists include botanical names in charts

<span style="color: green;">**Activity**</span>

restaurant tips dataset - https://raw.githubusercontent.com/balsaedi/RCourse_1/main/data/tips.csv

In this activity, you are required to use the restaurant tips dataset. The link to the dataset is provided above. Read the csv data to a data frame and plot the following graphs. 

  a. A bar chart to compare the average tips from male and female customers
  b. A histogram to show the distribution of bills 
  c. A scatter plot to show the relationship between bills and tips. 
    
```{r ex="activity-6a", type="sample-code"}
# Load the required libraries 
# CODE HERE

# Read the csv data to a data frame and plot the following graphs. 
# CODE HERE 

## a. A bar chart to compare the average tips from male and female customers
# CODE HERE

## b. A histogram to show the distribution of bills 
# CODE HERE

## c. A scatter plot to show the relationship between bills and tips. 
# CODE HERE
```

```{r ex="activity-6a", type="solution"}
# Load the required libraries 
library(ggplot2)

# Read the csv data to a data frame and plot the following graphs. 
data_link = "https://raw.githubusercontent.com/balsaedi/RCourse_1/main/data/tips.csv"
tips = read.csv(data_link)

## a. A bar chart to compare the average tips from male and female customers
# CODE HERE

## b. A histogram to show the distribution of bills 
# CODE HERE

## c. A scatter plot to show the relationship between bills and tips. 
# CODE HERE
```

## Working with inbuilt iris data 
We have worked with iris data set before. In this topic we shall ask some questions and plot them on a box plot to cement our understanding in `ggplot`

- **Plot a box plot to compare the sepal length in different iris species**
```{r}
graph = ggplot(iris, aes(x=Species, y=Sepal.Length))+ # the graph skeleton
  geom_boxplot()+aes(fill=Species) # aesthetics

print(graph)
```
Virginica have higher sepal length however some virginica have less sepal length than versicolor

Violin plot will show a better distribution than box plot
```{r}
graph = ggplot(iris, aes(x=Species, y=Sepal.Length))+
  geom_violin()+aes(fill=Species)

print(graph)
```

Since some data cleaning and manipulation is required before diving deep into visualizations, a package **tidyr** is introduced. It can be installed by invoking the command below to the console.  
```
install.packages("tidyr")
```

This package is a part of the tidyverse collection designed to make data cleaning and reshaping easy. It has multiple functions for instance `gather()`, `spread()`, `pivot_wider()`, `pivot_longer()`, `separate()`, `unite()` and `fill()`. In this section we shall not dive deep to explore tidyr package but we will borrow one of its important functions, `gather()` to manipulate iris data when plotting. This function, `gather()` is used to reshape data especially from a long format to a short format.

Invoke the library
```{r echo=TRUE, error=FALSE, warning=FALSE, message=FALSE}
library(tidyr) # can be used to convert from short format to long format
```

- Convert to a long format

`gather` command is used to convert data into fewer columns. The columns from sepal length to petal width are to be merged to a single column called measurements
```{r}
#Gather from sepal length to petal width column into type and measurement column 
iris_long = gather(iris, type, measurement, Sepal.Length:Petal.Width, factor_key = TRUE)
tail(iris_long)
```

Lets plot the data to a box plot 
```{r}
graph=ggplot(iris_long, aes(x=Species,y=measurement))+
  labs(x="Species", y="Length in cm", title = "variation in Petals and Sepals of Iris Species")+
  theme(axis.text = element_text(angle=90))+
  facet_grid(~type)+
  aes(fill=Species)+
  geom_boxplot()

print(graph)
```

the same can also be plotted to a violin plot 
```{r}
graph=ggplot(iris_long, aes(x=Species,y=measurement))+
  labs(x="Species", y="Length in cm", title = "variation in Petals and Sepals of Iris Species")+
  theme(axis.text = element_text(angle=90))+
  facet_grid(~type)+
  aes(fill=Species)+
  geom_violin()

print(graph)
```

<span style="color: green;">**Activity**</span>

In this activity, you will be required to use the famous R inbuilt data set, `mtcars`. Load the data set using the command, `data(mtcars)` and plot atleast two graphs

```{r ex="activity-6b", type="sample-code"}
# Load the ggplot2 library
# CODE HERE 

# Load the data 
# CODE HERE 

# Plot Chart 1
# CODE HERE 

# Plot Chart 2
# CODE HERE 
```

```{r ex="activity-6b", type="solution"}
# Load the ggplot2 library
# CODE HERE 

# Load the data 
# CODE HERE 

# Plot Chart 1
# CODE HERE 

# Plot Chart 2
# CODE HERE 
```

## CODE CHALLENGE
thyroid disease - https://www.kaggle.com/datasets/jainaru/thyroid-disease-data/data

HR analytics dataset - https://www.kaggle.com/datasets/pavansubhasht/ibm-hr-analytics-attrition-dataset/data

1. You will be required to retrieve the HR analytics employee attrition data set from kaggle. Click [here](https://www.kaggle.com/datasets/pavansubhasht/ibm-hr-analytics-attrition-dataset/data) to check out the data set. Read the data into an R data frame and answer the following questions using `read.csv` and `ggplot2` libraries. 

    a. Plot a histogram to show the distribution of age among the employees. 
    b. Draw a boxplot to show the employees' hourly rate in the sales department. Are there any outliers? What is the median employees' hourly rate from the boxplot? 
    c. How does age relate to monthly income? Are younger employees paid less? Plot a scatter plot to show the relationship between employees monthly income and age. **Note**:*Age should be in the x axis*
    d. How many male and female employees are their in the company? Are female employees more than male employees. Plot a bar chart to show. 
    e. Are male employees retained retained more than the female employees in the company. Filter to retrieve the retained employees("Attrition="No") only. 
    f. What is the marital status of the employees in the Sales department? Filter the data to retrieve the sales department then plot the value count of each marital status in the filtered data set on a pie chart. 
    g. Finally plot a heatmap to show the correlation of numeric columns. 
2. You will be required to download the thyroid disease csv file from [here](https://www.kaggle.com/datasets/jainaru/thyroid-disease-data/data). It contains 13 clinicopathologic features aiming to predict recurrence of well differentiated thyroid cancer. The data set was collected in duration of 15 years and each patient was followed for at least 10 years. The data was procured from thyroid disease datasets provided by the UCI Machine Learning Repository. You will load the csv file into R data frame, `read.csv`, and solve the following questions. 
    
    a. After the initial thyroid cancer treatment, has the disease recurred or not? Compare the two categories of patients (recurred and not recurred) using a bar chart.
    b. After physical examination of the thyroid gland and the surrounding structures, there were patients who had multinodular goiter, other single nodular gioters while rest of the patients had other complications. What are the composition of the patients? Plot a pie chart to show. 
    c. Plot a grouped bar chart to show the average risk and age of thyroid disease between males and female patient. Risk and Age columns against gender. 
    d. Plot a desnsity plot to show the distribution of age. Which age group is mostly affected by thyroid disease? 
    
