# Bioconductor

<script src="js/dcl.js"></script>
<link rel="stylesheet" href="https://maxcdn.bootstrapcdn.com/bootstrap/3.4.1/css/bootstrap.min.css">
<script src="https://maxcdn.bootstrapcdn.com/bootstrap/3.4.1/js/bootstrap.min.js"></script>

<script src="js/dcl.js"></script>
```{r include=FALSE}
tutorial::go_interactive(greedy=TRUE)
knitr::opts_chunk$set(echo = TRUE,error=TRUE)
```

Bioconductor is a repository for high-quality, peer-reviewed software packages for the analysis and comprehension of high-throughput genomic data. It contained specific tools that handle, analyze and visualize data from high-throughput assays for instance ChIP-Seq, single-cell RNA-Seq and RNA-Seq. It provides data structures such as `ExpressionSet`,`SummarizedExperiment`, and `GenomicRanges`that store and manipulate large scale genomic data. It also consists over 2000 packages just for data pre-processing, normalization, data analysis and visualization. Here are some of the popular packages in `BioConductor` has to offer;

- **`Biostrings`**: For efficient manipulation of large biological sequences.

- **`DESeq2`**: For differential gene expression analysis based on count data.

- **`GenomicRanges`**: For representing and manipulating genomic intervals and variables defined along a genome.

- **`AnnotationDbi`**: For handling and querying biological annotation databases.

- **`edgeR`**: Another popular package for differential expression analysis, particularly for RNA-Seq data.

## Installing BiocManager Package

To get started with Bioconductor, you need to install the `BiocManager` package, which facilitates the installation and management of Bioconductor packages. You can find detailed installation instructions in the [official documentation](https://cran.r-project.org/web/packages/BiocManager/vignettes/BiocManager.html).

Load the required libraries
```{r echo=TRUE, results='hide', error=FALSE, warning=FALSE, message=FALSE}
library(Biostrings)
```

After loading the necessary libraries, you can read sequence data into R. For instance, if you have sequence data stored in a multifasta file named multifasta_seq.txt, you can use the readDNAStringSet() function.
```{r}
seq = readDNAStringSet("data/multifasta_seq.txt")
seq
```

Each genomic sequence can be viewed separately by indexing 
```{r}
seq[[1]] #first genomic sequence
seq[[2]] # second genomic sequence
```

Convert to character to inspect its content.
```{r}
as.character(seq[[1]]) 
```

The `length` function from base R is used to count the number of genomic sequences. 
```{r}
length(seq)
```

- Names of the sequences
```{r}
names(seq)
```

- view the first 6 sequences
```{r}
head(seq)
```

- Checking the reverse compliment of sequence 
```{r}
seq_rc = reverseComplement(seq[[1]])
head(seq_rc)
tail(seq_rc)
as.character(seq_rc)
```

- Sequence composition -default - calculate the DNA composition
```{r}
composition=alphabetFrequency(seq)
print(composition)
```

- Specific composition - lets say we wanna look for A, T, G, C and N
```{r}
composition1=letterFrequency(seq, letters=c("A","T", "G", "C", "N"))
print(composition1)
```

Computing the GC Content 
```{r}
gc_content <- letterFrequency(seq, letters="GC", as.prob=TRUE)
gc_content
```

Calculate the mean of Gc content 
```{r}
mean(gc_content)
```

- Standard deviation 
```{r}
sd(gc_content)
```


```{r}
range(gc_content)
```

- Which GC has the maximum content 
```{r}
seq[which(gc_content==max(gc_content))]
```

## Searching and Analyzing the strings
A biologist can search for specific substrings within the sequences and analyze their positions and characteristics for instance;

- Search for substrings- in this case we search for `"GAATTC"` substring.
```{r}
ECoR1=vcountPattern("GAATTC", seq)
print(ECoR1)

HindIII = vcountPattern("AAGTT", seq)
print(HindIII)
```

-  Finding the position of substrings
```{r}
HindIII_pos=vmatchPattern("AAGCT", seq)
print(HindIII_pos)
```

- Translate 
```{r}
# translate(seq[[2]])
#Check functions available
methods(class = "DNAStringSet")
```

## Using seqinr
Another useful library in sequence analysis is `seqinr`. 
After installing the library by invoking `install.packages("seqinr")` to the console, load it to the workspace 
```{r echo=TRUE, error=FALSE, warning=FALSE, message=FALSE}
library(seqinr)
```

Reading the fasta file containing the protein sequence using the `read.fasta` function. 
```{r}
proseq = read.fasta(file="data/spike_multifasta.txt", seqtype="AA")
str(proseq)
```

Count the number of sequences 
```{r}
num_seq = length(proseq)
print(paste("Number of sequences = ", num_seq))
```

- Names of the sequences
```{r}
names=getName(proseq)
names 
```

- Annotation of sequences 
```{r}
annotation <- getAnnot(proseq)
annotation
```

- Length of individual sequences 
```{r}
len_seq = getLength(proseq)
len_seq
```

- Protein statistics 
```{r}
print(AAstat(proseq[[1]]))
```

It gives the type of proteins 
```{r}
for (i in 1:num_seq)
{
  print(names[i])
  print(AAstat(proseq[[i]]))
}
```

- Show commands in seqinr
```{r}
lseqinr()
```

## Using Short Read
This library can be installed by invoking the `BiocManager::install("ShortRead")` command to the console. 

load the library 
```{r echo=TRUE, error=FALSE, warning=FALSE, message=FALSE}
library(ShortRead)
```

- Reading the fastq sequences 
```{r}
fq = readFastq("data/file2.fastq")
fq
```

- Preview of first 3 sequences
```{r}
fq[1:3]
```

- Sequence information
```{r}
sread(fq)
```

Subset the sequence from second to fifth
```{r}
sread(fq[2:5])
```

Inspect the quality of the sequence 
```{r}
quality(fq)
```

- Number of sequence in the file 
```{r}
length(fq)
```

- Length of reads 
```{r}
readLength = width(fq)
readLength
```

- generating the summary of fasta sequences 
```{r}
fq_reads = sread(fq)

id_of_reads=id(fq)
id_of_reads[1:5]
fq_reads
```

- Seq length - we can generate the sequence lenght by converting it to a dataframe
```{r}
seq_length = as.data.frame(fq_reads@ranges@width)
seq_length
```

- Quality scores 
```{r}
quality_scores=quality(fq)
quality_scores[1:5]
```

Convert the scores to matrix and see as real numbers 
```{r}
scores=as(quality_scores, "matrix")
scores
```

- Mean score for each sequence 
```{r}
mean_score_for_each_sequence = rowMeans(scores, na.rm=TRUE)
mean_score_for_each_sequence
```

- Mean score for each position 
```{r}
mean_score_for_each_position = rowMeans(scores, na.rm=TRUE)
mean_score_for_each_position
```

Plot a boxplot for the scores 
```{r}
boxplot(scores, xlab="Position", ylab="Phred score")
```

## CODE CHALLENGE
In this code challenge you will be required to installed the Bioconductor locally as guided above. Thereafter, you will install `MultiAssayExperiment` package that is part of the Bioconductor project. Below is the code to install the `MultiAssayExperiment` locally;

```
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("MultiAssayExperiment")
```

You will load the package along with the ggplot2 library and work on the `genCov` data. `genCov` data is an inbuilt data set from `MultiAssayExperiment` that contains simulated data showcasing the integration and analysis of multiple types of data from different assays. Challenge yourself with the following activities to get a clear understanding of genomic coverage data;

1. Load the `genCov` using the code, `data("genCov", package = "MultiAssayExperiment")`.  
2. Check the structure `str()` of the data set and calculate the descriptive statistics.
3. Extract the coverage data with code `experiments(genCov)[["coverage"]]` and store in a variable `coverage_data`. Display the first 6 records of the the data. 
4. Extract and inspect the genomic ranges from `genCov` with `rowRanges()` command. Display the first 6 genomic ranges. 
5. Plot a line chart of the `"Coverage"` against the `"Genomic Position"`. 
6. Calculate the average coverage for each sample in the `coverageData` using `colMeans()` command. Display the result on a bar chart. 