---
title: "video_1"
output: html_document
date: "2024-04-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# R - Basic Concepts and applications 

<-- logo image for R --> 
**R (programming language)**

Language and environment for statistical computing and graphics 

<-- logo image for Rstudio --> 

Rstudio is an integrated development environment(IDE) for R. 

It includes a console, syntax-highlighting editor that supports direct code execution as well as tools for plotting history, debugging and workspace management. 
RStudio - installation link <-link in the first 7 minutes->

# Why R? 

1. **Developed by Statisticians** - R is Excellent for Statistical Computing and Analysis 
2. **Open-source (Free)** - R is an open-source programming language. Anyone can work with R without any license or fee. Due to this, R has a huge community that contributes to its environment. 
3. **A Large Variety of Specialized Libraries** - These libraries support and enhance the R development environment.
4. **R Studio** - A graphic user interface to R - easier to grasp and learn. 
5. **Cross-platform Support** - R is machine-independent. It supports the cross-platform operation. Thus, it is usable on many different operating systems.
6. **Can do Data Cleansing, Data Wrangling, and Web Scraping** - R can collect data from the internet through web scraping and other means. It can also perform data cleansing. R is also useful for data wrangling which is the process of converting raw data into the desired format for easier consumption. 
7. **Powerful Graphics** - R has extensive libraries that can produce production quality graphs and visualizations. 
8. **Compatible with other Programming Languages** - R is compatible with other languages like C, C++, and FORTRAN. Other languages like .NET, Java, Python can also directly manipulate objects.
9. Compatible with high performance computing and Big Data Analytics. 

# R - Look and Feel

<-- R screenshot image from the terminal-->

<-- default screenshot image for R Studio --> 
explain each section/quadrant
- text editor 
- stand input/output (console/terminal/jobs)
- all data and its properties quadrants 
- plots and file directory structure

# R studio
Outline the process of creating and running R script - a screenshot 
## Basic Mathematics like operators 

## Data types 
- Numeric
```{r}
a=23.5
class(a) #check the data type of a 
```

- Integers
```{r}
a=23L #add L to show it is an integer
class(a)
```

- Characters
```{r}
a="DNA"
class(a)
```

- Logical 
```{r}
a=TRUE #logical data types is either TRUE or FALSE only
class(a)
```

- Complex, 
```{r}
a=2+3i # Complex data types have 'i' at the end of each number
class(a)
```

- Raw
```{r}
a=charToRaw("DNA")
print(a)
class(a)

# convert back to character 
b=rawToChar(a)
class(b)
```

# Data Structure
Organization of one or multiple data values in a specific structures 

## Vector 
vector as “a single entity consisting of a collection of things.” 
```{r}
marks = c(23, 67, 98, 34, 98, 21)
print(marks)
marks

# what are the data types therein
class(marks)

#check if its a vector 
is.vector(marks)

#vector has no dimensions
dim(marks)

# length is used to count number of elements in vectors
length(marks)

# index is the position of an element in a vector 
# in R it starts at index 1 - lets say we find the third element by index 3
marks[3]

# access second to fourth values
marks[2:4]

# exclude value at specific index we use "-index" for instance 
marks[-2] # excludes value at index 2

# reverse the values of vector 
rev(marks)

# vector summary
summary(marks) #finds the summary/descriptive statistics

# ARITHMETIC WITH VECTORS 
## Calculate the total marks 
total_marks = sum(marks)
print(paste("Total marks=",total_marks))
```

<--End of video 1-->