---
title: "video_1"
output: html_document
date: "2024-04-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# R - Basic Concepts and applications 

<-- logo image for R --> 
**R (programming language)**

Language and environment for statistical computing and graphics 

<-- logo image for Rstudio --> 

Rstudio is an integrated development environment(IDE) for R. 

It includes a console, syntax-highlighting editor that supports direct code execution as well as tools for plotting history, debugging and workspace management. 
RStudio - installation link <-link in the first 7 minutes->

# Why R? 

1. **Developed by Statisticians** - R is Excellent for Statistical Computing and Analysis 
2. **Open-source (Free)** - R is an open-source programming language. Anyone can work with R without any license or fee. Due to this, R has a huge community that contributes to its environment. 
3. **A Large Variety of Specialized Libraries** - These libraries support and enhance the R development environment.
4. **R Studio** - A graphic user interface to R - easier to grasp and learn. 
5. **Cross-platform Support** - R is machine-independent. It supports the cross-platform operation. Thus, it is usable on many different operating systems.
6. **Can do Data Cleansing, Data Wrangling, and Web Scraping** - R can collect data from the internet through web scraping and other means. It can also perform data cleansing. R is also useful for data wrangling which is the process of converting raw data into the desired format for easier consumption. 
7. **Powerful Graphics** - R has extensive libraries that can produce production quality graphs and visualizations. 
8. **Compatible with other Programming Languages** - R is compatible with other languages like C, C++, and FORTRAN. Other languages like .NET, Java, Python can also directly manipulate objects.
9. Compatible with high performance computing and Big Data Analytics. 

# R - Look and Feel

<-- R screenshot image from the terminal-->

<-- default screenshot image for R Studio --> 
explain each section/quadrant
- text editor 
- stand input/output (console/terminal/jobs)
- all data and its properties quadrants 
- plots and file directory structure

# R studio
Outline the process of creating and running R script - a screenshot 
## Basic Mathematics like operators 

## Data types 
- Numeric
```{r}
a=23.5
class(a) #check the data type of a 
```

- Integers
```{r}
a=23L #add L to show it is an integer
class(a)
```

- Characters
```{r}
a="DNA"
class(a)
```

- Logical 
```{r}
a=TRUE #logical data types is either TRUE or FALSE only
class(a)
```

- Complex, 
```{r}
a=2+3i # Complex data types have 'i' at the end of each number
class(a)
```

- Raw
```{r}
a=charToRaw("DNA")
print(a)
class(a)

# convert back to character 
b=rawToChar(a)
class(b)
```

# Data Structure
Organization of one or multiple data values in a specific structures 

## Vector 
vector as “a single entity consisting of a collection of things.” 
```{r}
marks = c(23, 67, 98, 34, 98, 21)
print(marks)
marks

# what are the data types therein
class(marks)

#check if its a vector 
is.vector(marks)

#vector has no dimensions
dim(marks)

# length is used to count number of elements in vectors
length(marks)

# index is the position of an element in a vector 
# in R it starts at index 1 - lets say we find the third element by index 3
marks[3]

# access second to fourth values
marks[2:4]

# exclude value at specific index we use "-index" for instance 
marks[-2] # excludes value at index 2

# reverse the values of vector 
rev(marks)

# vector summary
summary(marks) #finds the summary/descriptive statistics

# ARITHMETIC WITH VECTORS 
## Calculate the total marks 
total_marks = sum(marks)
print(paste("Total marks=",total_marks))

## vector multiplication
double_marks =2 * marks 
marks
double_marks

## vector division 
half_marks = marks/2
half_marks

## multiply 2 vectors of equal lenght 
a = c(3, 5, 1)
b = c(7, 3, 9)
ab = a*b
ab
ba = b*a # is the same as ab
ba

## divide 2 vectors of equal length 
abdiv=a/b
abdiv

badiv=b/a
badiv

## multiplying vectors of unequal length
e=c(1,2,3,4,5)
f=c(1,2)
ef = e*f #it shows an error
ef #shows results since f is replicated to match e as f=c(1,2,1,2,1)

## Combining vectors 
a
b
z=c(a,b,a) #concatenates the vectors 
z

# Character Vectors 
my_name = c("My", "name", "is", "Vipin")
my_name[5] = "Singh" #insert at the end 
my_name
class(my_name)
summary(my_name)
print(paste(my_name, collapse=" ")) #combine the vector into a single string 

# Vectors with mixed data types
## vectors with numeric and characters 
numbers=c(1,"two", 3, "three")
numbers

class(numbers)

num1=numbers[1]
num1 

class(num1)
# num1+2 gives an error  
as.integer(num1) + 2 #need to convert to integer

## Named vectors
named_vector=c(EcoR1="GAATTC", HindIII="AAGCTT", Pst1="CTGCAG")
named_vector

names(named_vector)

named_vector["EcoR1"] # find the value of a vector by its name 

# Generating the number series 
series1=1:20
series1

series2=seq(1,30)
series2
series2[5] #access a value by index 
sum(series2) 

# generate numbers 0 to 10 incremented by 0.5
series3=seq(0, 10, by=0.5)
series3

# generate 10 numbers from 0 to 6
series4=seq(0, 6, length=10)
series4

# <--- add rep command for rep series---->

# NA data (Not available or blank)
marks=c(78,65, 98, 87, 89, NA)
sum(is.na(marks)) #Count the null values in a vector 





```

<--End of video 1-->